WHOOMP! (There it is):\\Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-12-07 ms.tex
Increment DCC version
Corrected proofs

\title[Rapid-response Bayesian sky localization]{WHOOMP! (There It Is) \\
Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients}
\author{Leo P. Singer}
\email{leo.singer@ligo.org}
\author{Larry R. Price}
\email{larryp@caltech.edu}
\affiliation{\acs{LIGO} Laboratory, California Institute of Technology, Pasadena, CA 91125, USA}
\pacs{04.80.Nn, 04.30.Tv, 02.50.Tt}

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect \acp{GW} from binary neutron star and neutron star\nobreakdashes--black hole mergers. These sources are also predicted to power a broad array of electromagnetic transients. Because the electromagnetic signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the gravitational wave observations. \ac{MCMC} methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-\ac{MCMC} sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search electromagnetic counterparts of compact binary mergers.
\end{abstract}

@@

The \acl{LIGO} (\acsu{LIGO}; \cite{AdvancedLIGOHarry,AdvancedLIGO}) has just begun taking data \cite{LIGOObservingScenarios} in its `Advanced' configuration. The two \ac{LIGO} detectors will ultimately increase their reach in volume within the local Universe by three orders of magnitude as compared to their initial configurations through 2010. They form the first parts of a sensitive global \ac{GW} detector network, soon to be augmented by Advanced Virgo \cite{AdvancedVirgo} and later by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India~\cite{LIGOIndia}.

The most readily detectable sources of \acp{GW} include binary neutron star mergers, with 0.4\nobreakdashes--400 events per year within the reach of Advanced LIGO at final design sensitivity~\cite{rates}. These binary systems are not only efficient sources of \acp{GW}, but also potential sources of detectable \ac{EM} transients from the aftermath of the tidal disruption of the \acp{NS}. \citet{MostPromisingEMCounterpart} argue that the most promising \ac{EM} counterparts are the hypothesized optical/infrared ``kilonovae'' powered by the radioactivate decay of $r$\nobreakdashes-process elements synthesized within the neutron\nobreakdashes-rich ejecta. These are expected to be faint, red, and peak rapidly, reaching an absolute magnitude of only $M_R \sim -13$ within a week, though rising several magnitudes brighter in the infrared~\citep{KilonovaHighOpacities}.

@@

Adding to the challenge of detecting a faint, short-lived optical transient, there is an extreme mismatch between the sky localization accuracy of \ac{GW} detector networks, $\sim$10\nobreakdashes--500\,deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, and the \acp{FOV} of 1\nobreakdashes--8\,m\nobreakdashes-class optical telescopes. Wide\nobreakdashes-field optical transient facilites such as BlackGEM~(0.6\,m/2.7\,deg$^2$), the \acl{ZTF}~(\acsu{ZTF}; 1.2\,m/47\,deg$^2$) \cite{ZTF}, the \acl{DECAM}~(\acsu{DECAM}; 4\,m/3\,deg$^2$), or the \acl{LSST}~(\acsu{LSST}; 8.4\,m/9.6\,deg$^2$), operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event. However, prompt and accurate \ac{GW} position reconstructions will be of the utmost importance for guiding the selection of fields to observe.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and electromagnetic emission from compact binaries. This involved several advances in the \ac{GW} data analysis \citep{CBCLowLatency}, including the first real\nobreakdashes-time matched\nobreakdashes-filter detection pipeline (\acs{MBTA}, \aclu{MBTA}; \cite{LIGOVirgoInspiralPipelineComparison}), a semi\nobreakdashes-coherent, ad hoc rapid triangulation code (Timing++), and the first version of a rigorous Bayesian \ac{MCMC} parameter estimation code (LALINFERENCE; \citealt{S6PE})---all in service of the first search for X\nobreakdashes-ray \citep{SwiftFollowup} and optical \citep{OpticalImageAnalysis} counterparts of \ac{GW} triggers, by a consortium of facilities. Despite the technical achievements in the \ac{GW} data analysis, there was an undesirable tradeoff between the speed as well as accuracy of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run~\cite{SiderySkyLocalizationComparison}.

The success of \ac{EM} follow\nobreakdashes-up of \ac{LIGO} events will depend critically on disseminating high quality sky localizations within a timescale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds to achieve approximately the same accuracy as the full \ac{MCMC} analysis. Our key insights are the following:

@@

    \item The Fisher information matrix will provide clues as to suitable forms of this likelihood function. Recent \ac{GW} parameter estimation literature has largely rejected the Fisher matrix,\footnote{Though not entirely; see \cite{EffectiveFisherMatrix}.} but this is mostly on the grounds of the abuse of the related \ac{CRLB} outside its realm of validity (i.e., at low to moderate \ac{SNR}) \cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, we recognize that the block structure of the Fisher matrix provides important insights, and is a useful quantity for checking the validity of the aforementioned likelihood function, quite independent of the \ac{CRLB}.

@@

    \item Thanks to a simple likelihood function and a well\nobreakdashes-characterized parameter space, we may dispense with costly and parallelization\nobreakdashes-resistant \ac{MCMC} integration, and instead perform the Bayesian marginalization with classic, deterministic, very low order Gaussian quadrature.

@@

We call this algorithm \ac{BAYESTAR}\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}. It is as fast as Timing++, but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} is providing localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a key role in enabling rapid follow-up observations.

This paper is organized as follows. In Section~\ref{sec:detection}, we describe the \ac{GW} signal model and sketch the standard detection algorithm, the matched filter bank. In Section~\ref{sec:parameter-estimation}, we describe Bayesian inference formalism and the prevailing method for inferring the parameters of detected candidates, \ac{MCMC} sampling. In Section~\ref{sec:bayestar-likelihood}, we propose the \ac{BAYESTAR} likelihood as a model for the uncertainty in the matched\nobreakdashes-filter parameter estimates, and discuss its relationship to and consistency with the likelihood for the full \ac{GW} data. In Section~\ref{sec:prior}, we describe the input to \ac{BAYESTAR} supplied by the detection pipeline, and the prior distribution on parameters. In Section~\ref{sec:marginal-posterior}, we explain the integration scheme by which the posterior probability distribution is calculated for a given sky location. In Section~\ref{sec:adaptive-sampling}, we show a scheme whereby the sky posterior is sampled on an adaptive \acs{HEALPix} grid. In Section~\ref{sec:run-time}, we report the running time of the algorithm on the hardware available on the \ac{LIGO} Data Grid. Finally, in Section~\ref{sec:case-study}, we quantify the sky localization performance on a comprehensive set of simulated \ac{GW} events.

@@

In the \ac{TD}, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}
    y_i(t) = x_i (t; \bm\theta) + n_i (t).
\end{equation}
%
In the \ac{FD},
%
\begin{equation}\label{eq:signal-model}
    Y_i (\omega) = \int_{-\infty}^\infty y(t) e^{-i \omega t} dt = X_i (\omega; \bm\theta) + N_i (\omega),
\end{equation}
%
where $X_i (\omega; \bm\theta)$ is the \ac{GW} signal given a parameter vector $\bm\theta$ that describes the \ac{GW} source, and $N_i (\omega)$ is that detector's Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|N_i(\omega)\right|^2\right] + E\left[\left|N_i(-\omega)\right|^2\right] = 2 E\left[\left|N_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf Y (\omega) \equiv \{Y_i (\omega)\}_i$.

Under the assumptions that the detector noise is Gaussian and that the noise from different detectors are uncorrelated, the likelihood of the observation, $\mathbf y$, conditioned upon the parameters $\bm\theta$, is a product of Gaussian distributions:
%
\begin{multline}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf Y; \bm\theta) = \prod_i p(Y_i | \bm\theta)
        \\
        \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|Y_i (\omega)
            - X_i(\omega; \bm\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{multline}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation, and intrinsic parameters describing the physical properties of the binary components\footnote{This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it may play a major role in the evolution and waveforms of rare close binaries formed by dynamical capture~\citep{PhysRevD.87.043004,2013PhRvD..87l7501H,2014PhRvD..90h4016H}, \ac{BNS} systems formed by binary stellar evolution should almost always circularize due to tidal interaction~\citep{0004-637X-572-1-407} and later \ac{GW} emission~\citep{PhysRev.136.B1224} long before the inspiral enters LIGO's frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\citep{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely: $\mathbf{S}_1 = \mathbf{S}_2 = 0$.}:
%
\begin{equation}\label{eq:params}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}

Assuming a nonprecessing circular orbit, we can write the \ac{GW} signal received by any detector as a linear combination of two basis waveforms, $H_0$ and $H_{\pi/2}$ \citep{PhysRevD.83.084002}. $H_0$ and $H_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the cosine and sine functions, being orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $H_0$ and $H_{\pi/2}$ are Fourier transforms of real functions, then $H_0(\omega) = H_0^*(-\omega)$ and $H_{\pi/2}(\omega) = H_{\pi/2}^*(-\omega)$, and we can write
%
\begin{equation}
    H_{\pi/2}(\omega) = H_0(\omega) \cdot
    \begin{cases}
        -i & \text{if } \omega \geq 0 \\
        i & \text{if } \omega < 0
    \end{cases}.
\end{equation}
%
For brevity, we define $H \equiv H_{0}$ and write all subsequent equations in terms of the $H$ basis vector alone. Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters, $\bm\theta_\mathrm{ex}$, into a few coefficients, and all dependence on the intrinsic parameters, $\bm\theta_\mathrm{in}$, into the basis waveform, by taking the Fourier transform of Equation~(2.8) of \cite{PhysRevD.83.084002}:
%
\begin{multline}\label{eq:full-signal-model}
    X_i(\omega; \bm\theta) = e^{-i \omega (t_\oplus - \mathbf{d}_i \cdot \mathbf{n})}
    \frac{r_{1,i}}{r}
    e^{2 i \phi_c}
    \\
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{\zeta\right\} - i
    \left(\cos\iota\right) \Im \left\{\zeta\right\}
    \right]
    H(\omega; \bm\theta_\mathrm{in})
\end{multline}
%
for $\omega \geq 0$, where
%
\begin{equation}
    \zeta = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}}^2 \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event and are presented in~\cite{ExcessPower}. In a coordinate system with the $x$ and $y$ axes aligned with the arms of a detector, the antenna pattern is given in spherical polar coordinates as
%
\begin{align}
    F_+ &= -\frac{1}{2}(1 + \cos^2 \theta) \cos{2\phi}, \\
    F_\times &= -\cos\theta \sin{2\phi}.
\end{align}
%
The unit vector $\mathbf{d}_i$ represents the position of detector $i$ in units of light travel time.\footnote{When considering transient \ac{GW} sources such as those that we are concerned with in this thesis, the origin of the coordinate system is usually taken to be the geocenter. For long\nobreakdashes-duration signals such as those from statically deformed neutron stars, the solar system barycenter is a more natural choice.} The vector $\mathbf{n}$ is the direction of the source. The negative sign in the dot product $-\mathbf{d}_i \cdot \mathbf{n}$ is present because the direction of travel of the \ac{GW} signal is opposite to that of its sky location. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register SNR=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms):
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|H(\omega; \boldsymbol \theta_\mathrm{in})\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

@@

We can predict the uncertainty in the detection pipeline's \ac{ML} estimates using the \ac{CRLB}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis to estimate parameter estimation uncertainty \cite{1996PhRvD..53.3033B,FairhurstTriangulation,2009PhRvD..79h4032A,WenLocalizationAdvancedLIGO,LIGOObservingScenarios,FairhurstLIGOIndia}\footnote{\label{footnote:template-banks}The Fisher matrix is also used in construction of \ac{CBC} matched filter banks. The common procedure is to place templates uniformly according to the determinant of the signal space metric, which is the Fisher matrix. This is equivalent to uniformly sampling the Jeffreys prior. In practice, this is done either by constructing a hexagonal lattice \citep{PhysRevD.76.102004} or sampling stochastically \citep{2009PhRvD..80j4014H,2009PhRvD..80b4009V,SBank,2010PhRvD..81b4004M,2014PhRvD..89b4003P}.}. As we noted, there are significant caveats to the \ac{CRLB} at low or moderate \ac{SNR} \cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, here we will be concerned more with gaining intuition from the block structure of the Fisher matrix than its numerical value. Furthermore, the Fisher matrix in its own right---independent of its suitability to describe the parameter covariance---is a well\nobreakdashes-defined property of any likelihood function, and we will exploit it as such in Section~\ref{sec:bayestar-likelihood}.

>>>>

\title[Rapid-response Bayesian sky localization]{WHOOMP! (There it is):\\Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients}
\author{Leo P. Singer}
\email{leo.singer@ligo.org}
\author{Larry R. Price}
\email{larryp@caltech.edu}
\affiliation{\acs{LIGO} Laboratory, California Institute of Technology, Pasadena, California 91125, USA}
\received{28 September 2015}
\pacs{04.80.Nn, 04.30.Tv, 02.50.Tt}

\begin{abstract}
Within the next few years, Advanced LIGO and Virgo should detect \aclp{GW} from \acl{BNS} and \acl{NSBH} mergers. These sources are also predicted to power a broad array of \acl{EM} transients. Because the \acl{EM} signatures can be faint and fade rapidly, observing them hinges on rapidly inferring the sky location from the \acl{GW} observations. \acl{MCMC} methods for gravitational\nobreakdashes-wave parameter estimation can take hours or more. We introduce BAYESTAR, a rapid, Bayesian, non-\acl{MCMC} sky localization algorithm that takes just seconds to produce probability sky maps that are comparable in accuracy to the full analysis. Prompt localizations from BAYESTAR will make it possible to search \acl{EM} counterparts of compact binary mergers.
\end{abstract}

@@

The Laser Interferometer Gravitational Wave Observatory (\acsu{LIGO})~\cite{AdvancedLIGOHarry,AdvancedLIGO}) has just begun taking data \cite{LIGOObservingScenarios} in its ``Advanced'' configuration. The two \ac{LIGO} detectors will ultimately increase their reach in volume within the local Universe by 3 orders of magnitude as compared to their initial configurations through 2010. They form the first parts of a sensitive global gravitational-wave (\acsu{GW}) detector network, soon to be augmented by Advanced Virgo~\cite{AdvancedVirgo} and later by the Japanese KAGRA facility~\cite{LCGT,KAGRAInterferometerDesign} and \ac{LIGO}\nobreakdashes--India~\cite{LIGOIndia}.

The most readily detectable sources of \acp{GW} include binary neutron star mergers, with 0.4\nobreakdashes--400 events per year within the reach of Advanced LIGO at its final design sensitivity~\cite{rates}. These binary systems are not only efficient sources of \acp{GW} but also potential sources of detectable \ac{EM} transients from the aftermath of the tidal disruption of the \acp{NS}. \citet{MostPromisingEMCounterpart} argue that the most promising \ac{EM} counterparts are the hypothesized optical/infrared ``kilonovae'' powered by the radioactive decay of $r$\nobreakdashes-process elements synthesized within the neutron\nobreakdashes-rich ejecta. These are expected to be faint and red and to peak rapidly, reaching an absolute magnitude of only $M_R \sim -13$ within a week, though rising several magnitudes brighter in the infrared~\citep{KilonovaHighOpacities}.

@@

Adding to the challenge of detecting a faint, short-lived optical transient, there is an extreme mismatch between the sky localization accuracy of \ac{GW} detector networks, $\sim$10\nobreakdashes--500\,deg$^2$ \citep{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, and the \acp{FOV} of 1\nobreakdashes--8\,m\nobreakdashes-class optical telescopes. Wide\nobreakdashes-field optical transient facilities such as BlackGEM~(0.6\,m/2.7\,deg$^2$), the \acl{ZTF}~(1.2\,m/47\,deg$^2$)~\cite{ZTF}, the \acl{DECAM}~(4\,m/3\,deg$^2$), or the \acl{LSST}~(8.4\,m/9.6\,deg$^2$), operated in ``target of opportunity'' mode, may be able to tile these large areas rapidly enough to find the one needle in the haystack that is connected to the \ac{GW} event. However, prompt and accurate \ac{GW} position reconstructions will be of the utmost importance for guiding the selection of fields to observe.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and \ac{EM} emission from compact binaries. This involved several advances in the \ac{GW} data analysis \citep{CBCLowLatency}, including the first real\nobreakdashes-time matched\nobreakdashes-filter detection pipeline, the \acl{MBTA}~\cite{LIGOVirgoInspiralPipelineComparison}; a semi\nobreakdashes-coherent, \textit{ad hoc} rapid triangulation code, Timing++; and the first version of a rigorous Bayesian \ac{MCMC} parameter estimation code, LALINFERENCE~\cite{S6PE}---all in service of the first search for X\nobreakdashes-ray \citep{SwiftFollowup} and optical \citep{OpticalImageAnalysis} counterparts of \ac{GW} triggers, by a consortium of facilities. Despite the technical achievements in the \ac{GW} data analysis, there was an undesirable tradeoff between the speed as well as accuracy of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run~\cite{SiderySkyLocalizationComparison}.

The success of \ac{EM} follow\nobreakdashes-up of \ac{LIGO} events will depend critically on disseminating high quality sky localizations within a time scale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds to achieve approximately the same accuracy as the full \ac{MCMC} analysis. Our key insights are the following:

@@

    \item The Fisher information matrix will provide clues as to suitable forms of this likelihood function. Recent \ac{GW} parameter estimation literature has largely rejected the Fisher matrix,\footnote{Though not entirely; see \cite{EffectiveFisherMatrix}.} but this is mostly on the grounds of the abuse of the related \ac{CRLB} outside its realm of validity of moderate to high \ac{SNR}~\cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, we recognize that the block structure of the Fisher matrix provides important insights and is a useful quantity for checking the validity of the aforementioned likelihood function, quite independent of the \ac{CRLB}.

@@

    \item Thanks to a simple likelihood function and a well\nobreakdashes-characterized parameter space, we may dispense with costly and parallelization\nobreakdashes-resistant \ac{MCMC} integration and instead perform the Bayesian marginalization with classic, deterministic, very low order Gaussian quadrature.

@@

We call this algorithm \ac{BAYESTAR}.\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces.}\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.} It is as fast as Timing++ but nearly as accurate as the rigorous full parameter estimation. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. Beginning with the first Advanced \acs{LIGO} observing run, \ac{BAYESTAR} is providing localizations within minutes of the detection of any \ac{BNS} merger candidate, playing a key role in enabling rapid follow-up observations.

This paper is organized as follows. In Sec.~\ref{sec:detection}, we describe the \ac{GW} signal model and sketch the standard detection algorithm, the matched\nobreakdashes-filter bank. In Sec.~\ref{sec:parameter-estimation}, we describe Bayesian inference formalism and the prevailing method for inferring the parameters of detected candidates, \ac{MCMC} sampling. In Sec.~\ref{sec:bayestar-likelihood}, we propose the \ac{BAYESTAR} likelihood as a model for the uncertainty in the matched\nobreakdashes-filter parameter estimates, and discuss its relationship to and consistency with the likelihood for the full \ac{GW} data. In Sec.~\ref{sec:prior}, we describe the input to \ac{BAYESTAR} supplied by the detection pipeline, and the prior distribution on parameters. In Sec.~\ref{sec:marginal-posterior}, we explain the integration scheme by which the posterior probability distribution is calculated for a given sky location. In Sec.~\ref{sec:adaptive-sampling}, we show a scheme whereby the sky posterior is sampled on an adaptive \ac{HEALPix} grid. In Sec.~\ref{sec:run-time}, we report the running time of the algorithm on the hardware available on the \ac{LIGO} Data Grid. Finally, in Sec.~\ref{sec:case-study}, we quantify the sky localization performance on a comprehensive set of simulated \ac{GW} events.

@@

In the \acl{TD}, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}
    y_i(t) = x_i (t; \bm\theta) + n_i (t).
\end{equation}
%
In the \acl{FD},
%
\begin{equation}\label{eq:signal-model}
    Y_i (\omega) = \int_{-\infty}^\infty y(t) e^{-i \omega t} dt = X_i (\omega; \bm\theta) + N_i (\omega),
\end{equation}
%
where $X_i (\omega; \bm\theta)$ is the \ac{GW} signal given a parameter vector $\bm\theta$ that describes the \ac{GW} source and $N_i (\omega)$ is that detector's Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|N_i(\omega)\right|^2\right] + E\left[\left|N_i(-\omega)\right|^2\right] = 2 E\left[\left|N_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf Y (\omega) \equiv \{Y_i (\omega)\}_i$.

Under the assumptions that the detector noise is Gaussian and that the noise from different detectors is uncorrelated, the likelihood of the observation, $\mathbf y$, conditioned upon the parameters $\bm\theta$, is a product of Gaussian distributions:
%
\begin{multline}\label{eq:gaussian-likelihood}
    \mathcal{L}(\mathbf Y; \bm\theta) = \prod_i p(Y_i | \bm\theta)
        \\
        \propto \exp \left[
        - \frac{1}{2} \sum_i \int_0^\infty \frac{\left|Y_i (\omega)
            - X_i(\omega; \bm\theta) \right|^2}{S_i(\omega)} \, d\omega
    \right].
\end{multline}

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components\footnote{This list of parameters involves some simplifying assumptions. Eccentricity is omitted; although it may play a major role in the evolution and waveforms of rare close binaries formed by dynamical capture~\citep{PhysRevD.87.043004,2013PhRvD..87l7501H,2014PhRvD..90h4016H}, \ac{BNS} systems formed by binary stellar evolution should almost always circularize due to tidal interaction~\citep{0004-637X-572-1-407} and later \ac{GW} emission~\citep{PhysRev.136.B1224} long before the inspiral enters LIGO's frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\citep{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely: $\mathbf{S}_1 = \mathbf{S}_2 = 0$.}:
%
\begin{equation}\label{eq:params}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}

Assuming a non\nobreakdashes-precessing circular orbit, we can write the \ac{GW} signal received by any detector as a linear combination of two basis waveforms, $H_0$ and $H_{\pi/2}$ \citep{PhysRevD.83.084002}. $H_0$ and $H_{\pi/2}$ are approximately ``in quadrature'' in the same sense as the cosine and sine functions, being orthogonal and out of phase by ${\pi/2}$ at all frequencies. If $H_0$ and $H_{\pi/2}$ are Fourier transforms of real functions, then $H_0(\omega) = H_0^*(-\omega)$ and $H_{\pi/2}(\omega) = H_{\pi/2}^*(-\omega)$, and we can write
%
\begin{equation}
    H_{\pi/2}(\omega) = H_0(\omega) \cdot
    \begin{cases}
        -i & \text{if } \omega \geq 0 \\
        i & \text{if } \omega < 0
    \end{cases}.
\end{equation}
%
For brevity, we define $H \equiv H_{0}$ and write all subsequent equations in terms of the $H$ basis vector alone. Then, we can write the signal model in a way that isolates all dependence on the extrinsic parameters, $\bm\theta_\mathrm{ex}$, into a few coefficients and all dependence on the intrinsic parameters, $\bm\theta_\mathrm{in}$, into the basis waveform, by taking the Fourier transform of Eq.~(2.8) of Ref.~\cite{PhysRevD.83.084002},
%
\begin{multline}\label{eq:full-signal-model}
    X_i(\omega; \bm\theta) = e^{-i \omega (t_\oplus - \mathbf{d}_i \cdot \mathbf{n})}
    \frac{r_{1,i}}{r}
    e^{2 i \phi_c}
    \\
    \left[
    \frac{1}{2} \left(1 + \cos^2 \iota\right) \Re \left\{\zeta\right\} - i
    \left(\cos\iota\right) \Im \left\{\zeta\right\}
    \right]
    H(\omega; \bm\theta_\mathrm{in})
\end{multline}
%
for $\omega \geq 0$, where
%
\begin{equation}
    \zeta = e^{-2 i \psi} \left(
    F_{+,i}(\alpha, \delta, t_\oplus) +
    i F_{\times,i}(\alpha, \delta, t_\oplus)
    \right).
\end{equation}
%
The quantities $F_{+,i}$ and $F_{\times,i}$ are the dimensionless detector antenna factors, defined such that $0 \leq {F_{+,i}}^2 + {F_{\times,i}}^2 \leq 1$. They depend on the orientation of detector $i$ as well as the sky location and sidereal time of the event and are presented in Ref.~\cite{ExcessPower}. In a coordinate system with the $x$ and $y$ axes aligned with the arms of a detector, the antenna pattern is given in spherical polar coordinates as
%
\begin{align}
    F_+ &= -\frac{1}{2}(1 + \cos^2 \theta) \cos{2\phi}, \\
    F_\times &= -\cos\theta \sin{2\phi}.
\end{align}
%
The unit vector $\mathbf{d}_i$ represents the position of detector $i$ in units of light travel time.\footnote{When considering transient \ac{GW} sources such as those that we are concerned with in this article, the origin of the coordinate system is usually taken to be the geocenter. For long\nobreakdashes-duration signals such as those from statically deformed neutron stars, the solar system barycenter is a more natural choice.} The vector $\mathbf{n}$ is the direction of the source. The negative sign in the dot product $-\mathbf{d}_i \cdot \mathbf{n}$ is present because the direction of travel of the \ac{GW} signal is opposite to that of its sky location. The quantity $r_{1,i}$ is a fiducial distance at which detector $i$ would register \ac{SNR}=1 for an optimally oriented binary (face\nobreakdashes-on, and in a direction perpendicular to the interferometer's arms):
%
\begin{equation}\label{eq:horizon}
r_{1,i} = 1 / \sigma_i, \qquad {\sigma_i}^2 = \int_0^\infty \frac{\left|H(\omega; \boldsymbol \theta_\mathrm{in})\right|^2}{S_i(\omega)} \,d\omega.
\end{equation}

@@

We can predict the uncertainty in the detection pipeline's \ac{ML} estimates using the \ac{CRLB}. The \ac{CRLB} has been widely applied in \ac{GW} data analysis to estimate parameter estimation uncertainty \cite{1996PhRvD..53.3033B,FairhurstTriangulation,2009PhRvD..79h4032A,WenLocalizationAdvancedLIGO,LIGOObservingScenarios,FairhurstLIGOIndia}.\footnote{\label{footnote:template-banks}The Fisher matrix is also used in construction of \ac{CBC} matched-filter banks. The common procedure is to place templates uniformly according to the determinant of the signal space metric, which is the Fisher matrix. This is equivalent to uniformly sampling the Jeffreys prior. In practice, this is done either by constructing a hexagonal lattice~\citep{PhysRevD.76.102004} or sampling stochastically~\citep{2009PhRvD..80j4014H,2009PhRvD..80b4009V,SBank,2010PhRvD..81b4004M,2014PhRvD..89b4003P}.} As we noted, there are significant caveats to the \ac{CRLB} at low or moderate \ac{SNR} \cite{BayesianBounds,UseAbuseFisherMatrix,FisherMatrixAsymptoticExpansions,InadequaciesFisherMatrix}. However, here we will be concerned more with gaining intuition from the block structure of the Fisher matrix than its numerical value. Furthermore, the Fisher matrix in its own right---independent of its suitability to describe the parameter covariance---is a well\nobreakdashes-defined property of any likelihood function, and we will exploit it as such in Sec.~\ref{sec:bayestar-likelihood}.
